---
title: "MB5370 Presentation Waln"
author: "Ryan Waln"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## INTRODUCTION

This document is used in conjunction with the associated data wrangling script 
to produce figures from the all_reefs data to visualize changes in hard coral cover, 
explore how these changes relate to environmental parameters, and if there are any resilience trends

Figures plotted include: 

- Fig 1: Earliest HC cover in the first year where ALL reefs have a measurement
- Fig 2: Current HC cover per reef site
- Fig 3: Dot-size proportional to avg HC change with color showing increase vs. decrease

```{r, message=FALSE, warning=FALSE, include=FALSE}
# R Setup & housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)

rm(list=ls())
objects()
```

```{r}
# Packages
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(dplyr)
library(ozmaps)    
```

## SOURCE SCRIPT FOR DATA WRANGLING
```{r}
source("../code/test.R")
```

## CONVERT SUMMARY TABLE TO sf SPATIAL OBJECT
```{r}
# use st_as_sf() to turn data frame into a spatial object to account for earth's curvature
# use CRS(4326 = WGS84) for Australia
# Note: Better to convert to spatial object on qmd file in order to play around with jittering for figures

all_reefs_sf <- st_as_sf(
  all_reefs_summary,
  coords = c("LONGITUDE", "LATITUDE"), # signify which columns have the coordinates 
  crs    = 4326                          # WGS84 geographic coordinate system
)

all_reefs_sf <- st_jitter(all_reefs_sf, amount = 0.1)  # add in jittering to reduce overlap of points

```

## LOAD QLD COASTLINE
```{r}
# Can use built in dataset to load in QLD coastline to lay sampling sites over

qld <- ozmap_states |>                         # All Aus. state polygons
  filter(NAME == "Queensland") |>              # keep only the QLD polygon
  st_transform(crs = 4326)                     # switch to WGS84 to match reef points
```

## DEFINE MAJOR COASTAL CITY LOCATIONS
```{r}
# Hard-code coordinates for the major cities along the GBR coast.
# Coordinates are approximate city-centre positions in WGS84.

cities <- data.frame(
  name      = c("Cairns", "Townsville", "Mackay", "Rockhampton"),
  longitude = c(145.77,    146.82,       149.18,   150.51),
  latitude  = c(-16.92,   -19.26,       -21.14,   -23.38)
) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # convert to sf point object
```

## SHARED MAP SETTINGS
```{r}

tmap_mode("plot")   # "plot" for saving figures; use "view" for interactive mode

# Bounding box derived from the reef points: used to frame all three maps
bbox_reefs <- st_bbox(all_reefs_sf)

# Common land polygon style applied to every figure
land_style <- list(
  col        = "lightgray",  # fill colour for the Queensland Coast
  border.col = "grey40",                # outline colour
  lwd        = 0.6                      # outline line width
)
```

## FIGURE 1 – STARTING HC COVER
```{r}

fig1 <- tm_shape(qld, bbox = bbox_reefs) +          # draw QLD coastline, cropped to reef extent
  tm_polygons(
    fill       = land_style$col,        # solid land fill color
    col        = land_style$border.col, # polygon border color
    lwd        = land_style$lwd         # border line width
  ) +
  tm_shape(cities) +                   # overlay city point locations
  tm_dots(
    fill = "black",                    # Display center of city as black dot
    size = 0.2                         # Set size of city dot
  ) +
  tm_shape(cities) +                   # repeat tm_shape so tm_text knows which layer to label
  tm_text(
    text     = "name",                 # set labels to use column with city names
    size     = 0.6,
    col      = "black",
    fontface = "bold",
    xmod     = -2.5,                   # negative value nudges label to the LEFT of the dot
    ymod     = -0.5                     # slight downward offset to sit beside rather than on the dot
  ) +
  tm_shape(all_reefs_sf) +             # overlay the reef point layer
  tm_symbols(
    shape       = 21,                   # R point shape 21: filled circle with border to represent sample sites
    fill        = "earliest_HC",        # color dot interior by earliest HC value
    fill.scale  = tm_scale_continuous(
      values = "brewer.rd_bu",       # color blind friendly palette (low to high cover)
      limits = c(0, 100)                # fix scale 0–100% so all figures are comparable
    ),
    fill.legend = tm_legend(
      title    = "2005 Hard Coral Cover (%)",
      position = c("right", "top")      # legend inside the plot, top-right corner
    ),
    col  = "black",                    # dot border (outline) color
    lwd  = 0.7,                         # dot border width
   
  ) +
  tm_scalebar(position = c("left", "bottom")) +         # add scale bar
  tm_compass(position  = c("right", "top"), size = 1) + # add compass
  tmap_style("natural")                                 # set style 

fig1   # display figure
```

##  FIGURE 2 – LATEST HC COVER
```{r}
fig2 <- tm_shape(qld, bbox = bbox_reefs) + # draw QLD coastline, cropped to reef extent
  tm_polygons(
    fill = land_style$col,              # solid land fill colour
    col  = land_style$border.col,       # polygon border colour
    lwd  = land_style$lwd               # border line width
  ) +
  tm_shape(cities) +                   # overlay city point locations
  tm_dots(fill = "black", size = 0.2) + # Display center of city as black dot 
  tm_shape(cities) +                   # repeat tm_shape so tm_text knows which layer to label
  tm_text(text = "name", size = 0.6, col = "black", fontface = "bold",
          xmod = -2.5, ymod = -0.5) +  # label placement next to dot
  tm_shape(all_reefs_sf) +            # overlay the reef point layer
  tm_symbols(
    shape       = 21,                   # filled circle with border
    fill        = "latest_HC",          # color dot interior by the most recent HC value
    fill.scale  = tm_scale_continuous(
      values = "brewer.rd_bu",        # color blind friendly palette (low to high cover)
      limits = c(0, 100)          # fix scale 0–100% so all figures are comparable
    ),
    fill.legend = tm_legend(
      title    = "2024 Hard Coral Cover (%)",
      position = c("right", "top")      # legend inside the plot, top-right corner
    ),
    col  = "black",                    # dot border (outline) color
    lwd  = 0.7,                         # dot border width
    
  ) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_compass(position  = c("right", "top"), size = 1) +
  tmap_style("natural")

fig2

```

#   FIGURE 3 – AVERAGE RATE OF CHANGE IN HC
```{r}
fig3 <- tm_shape(qld, bbox = bbox_reefs) +
  tm_polygons(
    fill = land_style$col,              # solid land fill colour
    col  = land_style$border.col,       # polygon border colour
    lwd  = land_style$lwd               # border line width
  ) +
  tm_shape(cities) +                   # overlay city point locations
  tm_dots(fill = "black", size = 0.2) + # Display center of city as black dot
  tm_shape(cities) +                   # repeat tm_shape so tm_text knows which layer to label
  tm_text(text = "name", size = 0.6, col = "black", fontface = "bold",
          xmod = -2.5, ymod = -0.5) +  # label placement next to dot
  tm_shape(all_reefs_sf) +
  tm_symbols(
    shape      = 21,                 # filled circle with border
    fill       = "avg_HC_change",
    fill.scale = tm_scale_continuous(
      values  = "brewer.rd_bu",   # Red (decrease), White (≈0), Blue (increase)
      midpoint = 0,                  # anchor the colour midpoint at zero change
      limits  = c(                   # symmetric limits centred on 0 for fair comparison
        -max(abs(all_reefs_summary$avg_HC_change), na.rm = TRUE),
        max(abs(all_reefs_summary$avg_HC_change), na.rm = TRUE)
      )
    ),
    fill.legend = tm_legend(
      title    = "Annual Avg % HC Change",
      position = c("right", "top")   # legend inside the plot, top-right corner
    ),
    col  = "black",                 # dot border (outline) color
    lwd  = 0.7,                      # dot border width
    size.legend = tm_legend(
      title    = "|Change| (annual %HC change )",
      position = c("right", "top")   # keep size legend in the same corner
    )
  ) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_compass(position  = c("right", "top"), size = 1) +
  tmap_style("natural")

fig3
```

#  SAVE FIGURES
```{r}

# Width = 8 inches, Height = 12 inches, DPI = 300 → publication-quality output

tmap_save(
  fig1,
  here("outputs/fig1_earliest_HC.png"),
  width = 8, height = 12, dpi = 300
)

tmap_save(
  fig2,
  here("outputs/fig2_latest_HC.png"),
  width = 8, height = 12, dpi = 300
)

tmap_save(
  fig3,
  here("outputs/fig3_HC_change.png"),
  width = 8, height = 12, dpi = 300
)
```