---
title: "MB5370 Presentation Waln"
author: "Ryan Waln"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## INTRODUCTION

This document is used in conjunction with the associated data wrangling script 
to produce figures from the all_reefs data to visualize changes in hard coral cover, 
and explore how these changes relate to the annual discharge of all rivers in NRM regions

Figures plotted include: 

- Fig 1: Map of sample sites with starting HC cover (first year where ALL reefs have a measurement)
- Fig 2: Map of sample sites with latest HC cover
- Fig 3: Map of sample sites with avg % HC change over course of survey data 
- Fig 4: Box plot of HC cover across NRM regions at start of survey
- Fig 5: Box plot of HC cover across NRM regions at end of survey
- Fig 6: Bar chart of average % HC change across NRM regions 
- Fig 7: Scatter plot of annual discharge vs HC by NRM region 

```{r, message=FALSE, warning=FALSE, include=FALSE}
# R Setup & housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)

rm(list=ls())
objects()
```

```{r}
# Packages
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(dplyr)
library(ozmaps)    
```

## SOURCE SCRIPT FOR DATA WRANGLING
```{r}
source("../code/MB5370_Presentation_Data_Wrangling.R")
```

## CONVERT SUMMARY TABLE TO sf SPATIAL OBJECT
```{r}
# use st_as_sf() to turn data frame into a spatial object to account for earth's curvature
# use CRS(4326 = WGS84) for Australia
# Note: Better to convert to spatial object on qmd file in order to play around with jittering for figures

all_reefs_sf <- st_as_sf(
  all_reefs_summary,
  coords = c("LONGITUDE", "LATITUDE"), # signify which columns have the coordinates 
  crs    = 4326                          # WGS84 geographic coordinate system
)

all_reefs_sf <- st_jitter(all_reefs_sf, amount = 0.05)  # add in jittering to reduce overlap of points

```

## LOAD QLD COASTLINE
```{r}
# Can use built in dataset to load in QLD coastline to lay sampling sites over

qld <- ozmap_states |>                         # all Aus. state polygons
  filter(NAME == "Queensland") |>              # keep only the QLD polygon
  st_transform(crs = 4326)                     # switch to WGS84 to match reef points
```

## DEFINE MAJOR COASTAL CITY LOCATIONS
```{r}
# Hard-code coordinates for the major cities along the GBR coast.
# Coordinates are approximate city-center positions in WGS84.

cities <- data.frame(
  name      = c("Cairns", "Townsville", "Mackay", "Rockhampton"),
  longitude = c(145.77,    146.82,       149.18,   150.51),
  latitude  = c(-16.92,   -19.26,       -21.14,   -23.38)
) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # convert to sf point object
```

## SHARED MAP SETTINGS
```{r}

tmap_mode("plot")   # "plot" for saving figures; use "view" for interactive mode

# Bounding box derived from the reef points: used to frame all three maps
bbox_reefs <- st_bbox(all_reefs_sf)

# Polygon style of coast applied to every figure
land_style <- list(
  col        = "lightgray",  # fill color for the Queensland Coast
  border.col = "grey40",                # outline color
  lwd        = 0.6                      # outline line width
)
```

## FIGURE 1 – MAP OF STARTING HC COVER
```{r}
fig1 <- tm_shape(qld, bbox = bbox_reefs) +          # draw QLD coastline, cropped to reef extent
  tm_polygons(
    fill       = land_style$col,        # solid land fill color
    col        = land_style$border.col, # polygon border color
    lwd        = land_style$lwd         # border line width
  ) +
  tm_shape(cities) +                   # overlay city point locations
  tm_dots(
    fill = "black",                    # Display center of city as black dot
    size = 0.2                         # Set size of city dot
  ) +
  tm_shape(cities) +                   # repeat tm_shape so tm_text knows which layer to label
  tm_text(
    text     = "name",                 # set labels to use column with city names
    size     = 0.6,
    col      = "black",
    fontface = "bold",
    xmod     = -2.5,                   # negative value nudges label to the LEFT of the dot
    ymod     = -0.5                     # slight downward offset to sit beside rather than on the dot
  ) +
  tm_shape(all_reefs_sf) +             # overlay the reef point layer
  tm_symbols(
    shape       = 21,                   # R point shape 21: filled circle with border to represent sample sites
    fill        = "earliest_HC",        # color dot interior by earliest HC value
    fill.scale  = tm_scale_continuous(
      values = "brewer.rd_bu",       # color blind friendly palette (low to high cover)
      limits = c(0, 100)                # fix scale 0–100% so all figures are comparable
    ),
    fill.legend = tm_legend(
      title    = "2005 Hard Coral Cover (%)",
      position = c("right", "top")      # legend inside the plot, top-right corner
    ),
    col  = "black",                    # dot border (outline) color
    lwd  = 0.7,                         # dot border width
   
  ) +
  tm_scalebar(position = c("left", "bottom")) +         # add scale bar
  tm_compass(position  = c("right", "top"), size = 1) + # add compass
  tmap_style("natural")                                 # set style 

fig1   # display figure
```

## FIGURE 2 – MAP OF LATEST HC COVER 
```{r}
fig2 <- tm_shape(qld, bbox = bbox_reefs) + # draw QLD coastline, cropped to reef extent
  tm_polygons(
    fill = land_style$col,              # solid land fill colour
    col  = land_style$border.col,       # polygon border colour
    lwd  = land_style$lwd               # border line width
  ) +
  tm_shape(cities) +                   # overlay city point locations
  tm_dots(fill = "black", size = 0.2) + # Display center of city as black dot 
  tm_shape(cities) +                   # repeat tm_shape so tm_text knows which layer to label
  tm_text(text = "name", size = 0.6, col = "black", fontface = "bold",
          xmod = -2.5, ymod = -0.5) +  # label placement next to dot
  tm_shape(all_reefs_sf) +            # overlay the reef point layer
  tm_symbols(
    shape       = 21,                   # filled circle with border
    fill        = "latest_HC",          # color dot interior by the most recent HC value
    fill.scale  = tm_scale_continuous(
      values = "brewer.rd_bu",        # color blind friendly palette (low to high cover)
      limits = c(0, 100)          # fix scale 0–100% so all figures are comparable
    ),
    fill.legend = tm_legend(
      title    = "2024 Hard Coral Cover (%)",
      position = c("right", "top")      # legend inside the plot, top-right corner
    ),
    col  = "black",                    # dot border (outline) color
    lwd  = 0.7,                         # dot border width
    
  ) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_compass(position  = c("right", "top"), size = 1) +
  tmap_style("natural")

fig2

```

## FIGURE 3 – MAP OF AVERAGE RATE OF CHANGE IN HC
```{r}
fig3 <- tm_shape(qld, bbox = bbox_reefs) +
  tm_polygons(
    fill = land_style$col,              # solid land fill color
    col  = land_style$border.col,       # polygon border color
    lwd  = land_style$lwd               # border line width
  ) +
  tm_shape(cities) +                   # overlay city point locations
  tm_dots(fill = "black", size = 0.2) + # display center of city as black dot
  tm_shape(cities) +                   # repeat tm_shape so tm_text knows which layer to label
  tm_text(text = "name", size = 0.6, col = "black", fontface = "bold",
          xmod = -2.5, ymod = -0.5) +  # label placement next to dot
  tm_shape(all_reefs_sf) +
  tm_symbols(
    shape      = 21,                 # filled circle with border
    fill       = "avg_HC_change",
    fill.scale = tm_scale_continuous(
      values  = "brewer.rd_bu",   # Red (decrease), White (≈0), Blue (increase)
      midpoint = 0,                  # anchor the color midpoint at zero change
      limits  = c(                   # symmetric limits centered on 0 for fair comparison
        -max(abs(all_reefs_summary$avg_HC_change), na.rm = TRUE),
        max(abs(all_reefs_summary$avg_HC_change), na.rm = TRUE)
      )
    ),
    fill.legend = tm_legend(
      title    = "Mean Annual % Change in HC",
      position = c("right", "top")   # legend inside the plot, top-right corner
    ),
    col  = "black",                 # dot border (outline) color
    lwd  = 0.7,                      # dot border width
  ) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_compass(position  = c("right", "top"), size = 1) +
  tmap_style("natural")

fig3
```
## FIGURE 4 – BOXPLOT OF STARTING HC COVER BY NRM REGION 
````{r}
fig4 <- ggplot(all_reefs_summary, aes(x = NRM_REGION,  # Plot earliest HC cover arranged by NRM region
                               y =  earliest_HC,
                               fill = NRM_REGION)) +           # color each region's box differently
  geom_boxplot(
    outlier.size  = 0.8,        # set size of outlier dots 
    outlier.alpha = 0.4,        # make outliers semi-transparent
    width         = 0.6         # set width of boxes
  ) +
  scale_fill_brewer(palette = "Set2") +   # colorblind-friendly palette
  scale_y_continuous(limits = c(0, 90))+   # set y axis from 0 to 90% to standardize range between 2 boxplots
  labs(
    x        = NULL,            # don't need x axis label with region labels
    y        = "Hard Coral Cover (%)",
    fill     = "NRM Region"
  ) +
  theme_bw() +                  # add grid lines 
  theme(
    legend.position = "none",   # don't need legend
    axis.text.y     = element_text(size = 10)   # edit text for region labels
  )

fig4   
````

## FIGURE 5 – BOXPLOT OF LATEST HC COVER BY NRM REGION 
```{r}
fig5 <- ggplot(all_reefs_summary, aes(x = NRM_REGION,  # Plot latest HC cover arranged by NRM region
                               y =  latest_HC,
                               fill = NRM_REGION)) +           # color each region's box differently
  geom_boxplot(
    outlier.size  = 0.8,        # set size of outlier dots 
    outlier.alpha = 0.4,        # make outliers semi-transparent
    width         = 0.6         # set width of boxes
  ) +
  scale_fill_brewer(palette = "Set2") +   # colorblind-friendly palette
  scale_y_continuous(limits = c(0, 90))+   # set y axis from 0 to 80%
  labs(
    x        = NULL,            # don't need x axis label with region labels
    y        = "Hard Coral Cover (%)",
    fill     = "NRM Region"
  ) +
  theme_bw() +                  # add grid lines 
  theme(
    legend.position = "none",   # don't need legend
    axis.text.y     = element_text(size = 10)   # edit text for region labels
  )

fig5   
```

## FIGURE 6 – BAR CHART OF AVERAGE CHANGE IN HC BY NRM REGION
```{r}
# Summarize the per-site avg_HC_change values from all_reefs_summary
# into a single mean ± SE for each NRM region.
fig6_data <- all_reefs_summary |>
  group_by(NRM_REGION) |>
  summarise(
    mean_change = mean(avg_HC_change),            # mean annual HC change across all sites in region
    se_change   = sd(avg_HC_change) / sqrt(n()),  # standard error for error bars
    n_sites     = n(),                             # record sample size for label
    .groups     = "drop"
  ) |>
  mutate(
    # Tag each region as net gain or net loss for color coding
    direction = if_else(mean_change >= 0, "Increase", "Decrease")
  )

fig6 <- ggplot(fig6_data,
               aes(x     = reorder(NRM_REGION, mean_change),  # order regions from largest to smallest decrease for visual design
                   y     = mean_change,
                   fill  = direction)) +                       # color bars by direction of change
  geom_col(width = 0.65) +                  # geom_col uses the y value directly (not counts)
  geom_errorbar(
    aes(ymin = mean_change - se_change,     # lower bound of error bar
        ymax = mean_change + se_change),    # upper bound of error bar
    width = 0.25,                           # set width of the error bars
    colour = "grey30"                       # color error bars
  ) +
  geom_hline(yintercept = 0,               # reference line at zero to clearly separate gain from loss
             linetype = "dashed",
             colour   = "black",
             linewidth = 0.5) +
  geom_text(
    aes(label = paste0("n = ", n_sites)),   # annotate each bar with its sample size
    hjust  = -0.5,  # Move label horizontally
    vjust  = 1.8,   # Move label vertically 
    size   = 3.2,
    colour = "grey20"
  ) +
 
  scale_fill_manual(
    values = c("Increase" = "#2166AC",      # blue for positive change
               "Decrease" = "#D6604D")      # red for negative change
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.15, 0.2)) # adds padding so the n= labels don't get clipped
  ) +
  labs(
    
    x        = NULL,                        # don't need x axis labels
    y        = "Mean Annual % Change in HC (± SE)",
    fill     = NULL                         # don't need coloring between NRM regions 
  ) +
  theme_bw() +
  theme(
    legend.position = "none",             # don't need legend
    axis.text.y     = element_text(size = 10)
  )

fig6  
```

## FIGURE 7 – SCATTER PLOT OF ANNUAL DISCHARGE vs HC COVER BY NRM REGION
````{r}
# Each point is one NMR-year observation.
# A linear regression line is fitted per NRM region (using geom_smooth)
# to show whether higher discharge years correspond to lower HC cover within each region.
# facet_wrap splits the plot into one panel per NRM region for easier comparison.

fig7 <- ggplot(discharge_HC,
               aes(x = total_discharge,   # x-axis: annual discharge (Mega Liters)
                   y = mean_HC)) +         # y-axis: mean HC cover (%) in that NRM region-year
  geom_point(
    aes(colour = NRM_REGION),             # color points by region
    alpha = 0.6,                           # semi-transparent so overlapping points are visible
    size  = 2
  ) +
  geom_smooth(
    method  = "lm",                        # fit a linear regression line per panel to find any trends
    se      = TRUE,                        # show 95% confidence interval ribbon around the line
    colour  = "grey20",linewidth = 0.8     # set line color and width 
  ) +
  facet_wrap(~ NRM_REGION,                # one panel per NRM region
             scales = "free_x") +          # put x-axis scale on each panel
  
  scale_colour_brewer(palette = "Set2") +  # colorblind-friendly palette matching other figures
  scale_x_continuous(
    labels = scales::label_comma()         # add commas to large discharge numbers 
  ) +
  labs(
    title  = "Annual River Discharge vs Hard Coral Cover by NRM Region",
    subtitle = "Each point = one subregion × year; line = linear regression with 95% CI",
    x      = "Total Annual Discharge (ML)",
    y      = "Mean Hard Coral Cover (%)"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",             # region label is already on each facet panel
    strip.text      = element_text(face = "bold", size = 10),  # bold facet panel titles
    axis.text.x     = element_text(angle = 30, hjust = 1)      # angle x-axis labels to prevent overlap
  )

fig7
````


## STATISTICAL ANALYSIS – DISCHARGE vs HC COVER BY NRM REGION
```{r}
# Run a simple linear regression on mean_HC vs total_discharge
# for each NRM region and extract the key statistics:
#   - Pearson correlation coefficient (r)
#   - regression slope (β) and its p-value
#   - R² (proportion of variance in HC explained by discharge)
# This gives a formal test of whether discharge predicts HC cover within each region.

discharge_stats <- discharge_HC |>
  group_by(NRM_REGION) |>
  summarise(
    n          = n(),   # number of NRM-year observations
    
    # Run Pearson correlation
    r          = cor(total_discharge, mean_HC,                 
                     use = "complete.obs"),
    r_squared  = summary(lm(mean_HC ~ total_discharge))$r.squared,   # Extract R² from linear model
    slope      = coef(lm(mean_HC ~ total_discharge))[2],              #Extract regression slope (% HC per ML)
    p_value    = summary(lm(mean_HC ~ total_discharge))$coefficients[2, 4],  #Extract p-value for slope
    .groups    = "drop"
  ) |>
  mutate(
    # Flag significance at 0.05 
    significant = if_else(p_value < 0.05, "Yes", "No"),
    # Round values for clean display
    across(c(r, r_squared, slope), ~ round(.x, 3)),
    p_value     = round(p_value, 4)
  )

# Note: a negative slope means higher discharge years correspond to lower HC cover.
print(discharge_stats)
```

## SAVE FIGURES
```{r}
# Width = 8 inches, Height = 12 inches, DPI = 300 (publication-quality output)

tmap_save(
  fig1,
  here("outputs/fig1_earliest_HC.png"),
  width = 8, height = 12, dpi = 300
)

tmap_save(
  fig2,
  here("outputs/fig2_latest_HC.png"),
  width = 8, height = 12, dpi = 300
)

tmap_save(
  fig3,
  here("outputs/fig3_HC_change.png"),
  width = 8, height = 12, dpi = 300
)

ggsave(
  here("outputs/fig4_STARTING_HC_boxplot_by_NRM.png"),
  plot   = fig4,
  width  = 10,
  height = 6,
  dpi    = 300
)

ggsave(
  here("outputs/fig5_END_HC_boxplot_by_NRM.png"),
  plot   = fig5,
  width  = 10,
  height = 6,
  dpi    = 300
)

ggsave(
  here("outputs/fig6_avg_HC_change_by_NRM.png"),
  plot   = fig6,
  width  = 10,
  height = 6,
  dpi    = 300
)

ggsave(
  here("outputs/fig7_discharge_vs_HC_by_NRM.png"),
  plot   = fig7,
  width  = 12,
  height = 8,
  dpi    = 300
)
```
